---
title: "Clustering"
author: "Victoria Muckerson"
date: "8/9/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## This code applies hierarchical clustering methods to the most variable genes present in three glomerular data sets. The data sets consist of various chronic kidney diseases and the results are visualized with a heatmap. The final product is a pca plot of the most variable genes so it is possible to compare the unsupervised clustering with the underlying molecular patterns.

## The libraries being used must be uploaded
``` {r}
library(cluster)
library(dbscan)
library(BiocManager)
library(cluster)
library(mclust)
library(tidyverse)
library(ConsensusClusterPlus)
library(ALL)
library(dplyr)
library(ComplexHeatmap)
library(formattable)
library(ggbiplot)
```

## The data being used was previously created using the "Data_upload_and_normalization" script in github here:
https://github.com/vmuckerson/Chronic_kidney_disease/blob/master/Data_upload_and_normalization.R

```{r}
setwd("/Users/saezlab/Documents/CKD_Data/Glom_Data/")


data1 <- readRDS(file = "./all_104948.rds")
pca1 <- readRDS(file = "./pca_104948.rds")
Pdata1 <- readRDS(file = "./Pdata_104948.rds")
scdata1 <- scale(data1)

data2 <- readRDS(file = "./all_GSE37460.rds")
pca2 <- readRDS(file = "./pca_GSE37460.rds")
Pdata2 <- readRDS(file = "./Pdata_GSE37460.rds")
scdata2 <- scale(data2)

data3 <- readRDS(file = "./all_GSE32591.rds")
pca3 <- readRDS(file = "./pca_GSE32591.rds")
Pdata3 <- readRDS(file = "./Pdata_GSE32591.rds")
scdata3 <- scale(data3)
```


## The most variable genes in each data set were selected, sorted, and used to index the datasets for clustering.

```{r}
#Selecting most variable genes
var1 <- rev(sort(apply(data1, 1, FUN = var)))
var2 <- rev(sort(apply(data2, 1, FUN = var)))
var3 <- rev(sort(apply(data3, 1, FUN = var)))

top1 <- var1[1:2000]
top2 <- var2[1:2000]
top3 <- var3[1:2000]

data1 <- scale(data1[rownames(data1) %in% names(top1),])
data2 <- scale(data2[rownames(data2) %in% names(top2),])
data3 <- scale(data3[rownames(data3) %in% names(top3),])
```


#The most variable genes are then clustered using hierarchical clustering techniques driven by pearson distances


```{r}

#Consensus Clustering------------

mad1 <- apply(data1, 1, mad)
d1 <- data1[rev(order(mad1))[1:1000],]
d1 <- sweep(d1,1,apply(d1,1,median,na.rm=TRUE))
d1 <- na.omit(d1)
results1 <- ConsensusClusterPlus(d1, maxK=6,
                                 reps=1000,
                                 pItem = 0.8,
                                 pFeature=1,
                                 title="Consensus Cluster on GSE104948 most variable genes",
                                 innerLinkage="average",
                                 finalLinkage = "average",
                                 clusterAlg="hc",
                                 distance="pearson")
icl1 <- calcICL(results1)



mad2 <- apply(data2, 1, mad)
d2 <- data2[rev(order(mad2))[1:5000],]
d2 <- sweep(d2,1,apply(d2,1,median,na.rm=TRUE))
d2 <- na.omit(d2)
results2 <- ConsensusClusterPlus(d2, maxK=6,
                                 reps=1000,
                                 pItem = 0.8,
                                 pFeature=1,
                                 title="Consensus Cluster on GSE37460 most variable genes",
                                 innerLinkage="average",
                                 finalLinkage = "average",
                                 clusterAlg="hc",
                                 distance="pearson")
icl2 <- calcICL(results2)



mad3 <- apply(data3, 1, mad)
d3 <- data3[rev(order(mad3))[1:5000],]
d3 <- sweep(d3,1,apply(d3,1,median,na.rm=TRUE))
d3 <- na.omit(d3)
results3 <- ConsensusClusterPlus(d3, maxK=6,
                                 reps=1000,
                                 pItem = 0.8,
                                 pFeature=1,
                                 title="Consensus Cluster on GSE32591 most variable genes",
                                 innerLinkage="average",
                                 finalLinkage = "average",
                                 clusterAlg="hc",
                                 distance="pearson")
icl3 <- calcICL(results3)
```



## The results of the clustering are used to create a heatmap that is annotated with the cluster assignment and the disease diagnosis status


```{r}
#CLUSTERING HEAT MAPS------------------------------------------

mat1 <- results1[[4]]$consensusMatrix
df1 <- data.frame(Class = Pdata1$`diagnosis:ch1`,
                  Consensus_class=as.character(results1[[4]]$consensusClass))

h1 <- HeatmapAnnotation(df = df1)
dimnames(mat1) <- list(colnames(Pdata1$`diagnosis:ch1`), colnames(Pdata1$`diagnosis:ch1`))
Heatmap(mat1,
        top_annotation = h1,
        show_column_names = FALSE,
        show_row_names = FALSE,
        column_title = "Consensus Clustering of GSE104948 most variable genes")



mat2 <- results2[[4]]$consensusMatrix
df2 <- data.frame(Class = Pdata2$disease,
                  Consensus_class=as.character(results2[[4]]$consensusClass))

h2 <- HeatmapAnnotation(df = df2)
dimnames(mat2) <- list(colnames(Pdata2$disease), colnames(Pdata2$disease))
Heatmap(mat2,
        top_annotation = h2,
        show_column_names = FALSE,
        show_row_names = FALSE,
        column_title = "Consensus Clustering of GSE37460 most variable genes")



mat3 <- results3[[4]]$consensusMatrix
df3 <- data.frame(Class = Pdata3$`disease status:ch1`,
                  Consensus_class=as.character(results3[[4]]$consensusClass))

h3 <- HeatmapAnnotation(df = df3)
dimnames(mat3) <- list(colnames(Pdata3$`disease status:ch1`), colnames(Pdata3$`disease status:ch1`))
Heatmap(mat3,
        top_annotation = h3,
        show_column_names = FALSE,
        show_row_names = FALSE,
        column_title = "Consensus Clustering of GSE32591 most variable genes")
```



## Finally a pca plot for each data sets most variable genes was created to visually compare the unsupervised hierarchical clustering results with the underlying molecular patterns in the data 


```{r}

###PCA using most variable genes------------------------------------
matdata1 <- as.matrix(data1)
pca1<- prcomp(t(matdata1), center=TRUE, scale. = TRUE)
ggbiplot(pca1,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata1$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104948")



matdata2 <- as.matrix(data2)
pca2<- prcomp(t(matdata2), center=TRUE, scale. = TRUE)
ggbiplot(pca2,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata2$disease) +
  ggtitle("PCA of GSE37460")



matdata3 <- as.matrix(data3)
pca3<- prcomp(t(matdata3), center=TRUE, scale. = TRUE)
ggbiplot(pca3,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata3$disease) +
  ggtitle("PCA of GSE32591")
```


## The unspecific disease grouping in the pcas confirms that the results from the clustering methods is not out of the ordinary. However, the difficulty grouping/clustering by molecular disease expression makes further analysis difficult. Therefore, in an attempt to further divide the diseases, a differential expression analysis was performed here:

https://github.com/vmuckerson/Chronic_kidney_disease/blob/master/Diff_expression_analysis.R

