---
title: "Clustering"
author: "Victoria Muckerson"
date: "8/9/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code applies hierarchical clustering methods to the most variable genes
present in three glomerular data sets. The data sets consist of various chronic
kidney diseases and the results are visualized with a heatmap. The final product
is a pca plot of the most variable genes so it is possible to compare the
unsupervised clustering with the underlying molecular patterns.

## Load Libraries
``` {r}

library(cluster)
library(dbscan)
library(BiocManager)
library(cluster)
library(mclust)
library(tidyverse)
library(ConsensusClusterPlus)
library(ALL)
library(dplyr)
library(ComplexHeatmap)
library(formattable)
library(ggbiplot)
```

The data being used was previously created using the
"Data_upload_and_normalization" script in github here:
https://github.com/vmuckerson/Chronic_kidney_disease/blob/master/Data_upload_and_normalization.R

```{r}
setwd("/Users/saezlab/Documents/CKD_Data/Glom_Data/")

data1 <- readRDS(file = "./all_104948.rds")
pca1 <- readRDS(file = "./pca_104948.rds")
Pdata1 <- readRDS(file = "./Pdata_104948.rds")
scdata1 <- scale(data1)

data2 <- readRDS(file = "./all_GSE37460.rds")
pca2 <- readRDS(file = "./pca_GSE37460.rds")
Pdata2 <- readRDS(file = "./Pdata_GSE37460.rds")
scdata2 <- scale(data2)

data3 <- readRDS(file = "./all_GSE32591.rds")
pca3 <- readRDS(file = "./pca_GSE32591.rds")
Pdata3 <- readRDS(file = "./Pdata_GSE32591.rds")
scdata3 <- scale(data3)
```

#Select most variable genes
The most variable genes in each data set were selected, sorted, and used to 
index the datasets for clustering.

```{r}
#Selecting most variable genes
datalist <- list(data1, data2, data3)

datalist <- list(data1, data2, data3)
pdatalist <- list(Pdata1, Pdata2, Pdata3)

topgenes <- datalist %>% lapply(FUN = function(data){
                            x <- rev(sort(apply(data, 1, FUN = var)))
                            return(x[1:2000])
                          })

topdata <- datalist %>% map2(topgenes,~scale(.x[rownames(.x) %in% names(.y),]))

```


#Clustering
The most variable genes are then clustered using hierarchical clustering 
techniques driven by pearson distances


```{r}

#Consensus Clustering------------

madlist <- topdata %>% lapply(FUN = function(data){
  apply(data, 1, mad)
})
d1 <- topdata %>% map2(madlist, ~.x[rev(order(.y))[1:1000],])
d1 <- d1 %>% lapply(FUN = function(data){
  sweep(data, 1, apply(data, 1, median, na.rm=TRUE))
})
d1 <- na.omit(d1)


results <- d1 %>% lapply(FUN = function(data){
  ConsensusClusterPlus(as.matrix(data),
                       maxK=6,
                       reps=1000,
                       pItem = 0.8,
                       pFeature=1,
                       innerLinkage="average",
                       finalLinkage = "average",
                       clusterAlg="hc",
                       distance="pearson")
})
  
icl <- lapply(resultslist, calcICL)

```


#Heatmap of Clustering
The results of the clustering are used to create a heatmap that is annotated
with the cluster assignment and the disease diagnosis status


```{r}
#CLUSTERING HEAT MAPS------------------------------------------

mats <- results %>% lapply(FUN = function(data){
  data[[6]]$consensusMatrix
})
Class <- lapply(pdatalist, FUN = function(data){
  data$`diagnosis:ch1`
})
Cluster <- as.character(lapply(results, FUN = function(data){
  data[[4]]$consensusClass
}))

df1 <- data.frame(Phenotype = Class[[1]],
                  Consensus_class = Cluster[[1]])

h1 <- HeatmapAnnotation(df = df1)
dimnames(mats[[1]]) <- list(colnames(pdatalist[[1]]$`diagnosis:ch1`),
                            colnames(pdatalist[[1]]$`diagnosis:ch1`))
Heatmap(mat1,
        top_annotation = h1,
        show_column_names = FALSE,
        show_row_names = FALSE,
        column_title = "Consensus Clustering of GSE104948 most variable genes")



df2 <- data.frame(Phenotype = Class[[2]],
                  Consensus_class = Cluster[[2]])

h2 <- HeatmapAnnotation(df = df2)
dimnames(mats[[2]]) <- list(colnames(pdatalist[[2]]$disease),
                            colnames(pdatalist[[2]]$disease))
Heatmap(mat2,
        top_annotation = h2,
        show_column_names = FALSE,
        show_row_names = FALSE,
        column_title = "Consensus Clustering of GSE37460 most variable genes")



df3 <- data.frame(Phenotype = Class[[3]],
                  Consensus_class = Cluster[[3]])

h3 <- HeatmapAnnotation(df = df3)
dimnames(mats[[3]]) <- list(colnames(pdatalist[[3]]$`diagnosis:ch1`),
                            colnames(pdatalist[[3]]$`diagnosis:ch1`))
Heatmap(mat3,
        top_annotation = h3,
        show_column_names = FALSE,
        show_row_names = FALSE,
        column_title = "Consensus Clustering of GSE32591 most variable genes")
```



##PCA of most variable genes
Finally a pca plot for each data sets most variable genes was created to
visually compare the unsupervised hierarchical clustering results with the
underlying molecular patterns in the data.


```{r}

###PCA using most variable genes------------------------------------
matdata1 <- as.matrix(data1)
pca1<- prcomp(t(matdata1), center=TRUE, scale. = TRUE)
ggbiplot(pca1,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata1$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104948")



matdata2 <- as.matrix(data2)
pca2<- prcomp(t(matdata2), center=TRUE, scale. = TRUE)
ggbiplot(pca2,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata2$disease) +
  ggtitle("PCA of GSE37460")



matdata3 <- as.matrix(data3)
pca3<- prcomp(t(matdata3), center=TRUE, scale. = TRUE)
ggbiplot(pca3,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata3$disease) +
  ggtitle("PCA of GSE32591")
```


The unspecific disease grouping in the pcas suggests that the poorly clustered
results from the clustering methods is to be expected. However, the difficulty
grouping/clustering by molecular disease expression makes further analysis
difficult. Therefore, in an attempt to further divide the diseases, a
differential expression analysis was performed here:
https://github.com/vmuckerson/Chronic_kidney_disease/blob/master/Diff_expression_analysis.R

A more functional analysis on this data was performed here:
https://github.com/saezlab/VM_CKD_saezlab/blob/master/Functional_Analysis.Rmd


