---
title: "Data download and visualization"
author: Victoria Muckerson
date: 8/8/2019
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = FALSE)
```

## Summary

This document outlines the uploading of a chronic kidney disease data set 
GSE104948 (glomerular tissue) from GEO. The data set contains data obtained via 
two different platforms (GPL22945 and GPL24120); for batch effect minimization 
when comparing to other datasets, only the platform GPL24120 is used for 
analysis. This document concludes with the production of pca and tsne plots 
colored by a phenotypic variable.


## Download necessary libraries

```{r}

#Download libraries

library(GEOquery)
library(qusage)
library(Biobase)
library(annotate)
library(hgu133a.db)
library(ggbiplot)
library(tidyverse)
library(Rtsne)

```

## Downloading the data
The data is downloaded from GEO.

```{r}

#Download the preprocessed data from GEO

mydata <- getGEO(GEO = 'GSE104948')
findata <- mydata[[2]]
otherplat <- mydata[[1]]
```


##Download phenotypic, expression, and feature data 
Though only one platform is used for analysis, the feature data annotation for the data from the other platform contains necessary information 
that the data of interest does not. Therefore, here, the feature data for both
platforms is downloaded.

```{r}
#Phenotypic

Pdata <- pData(findata)
Pdata$`diagnosis:ch1`[is.na(Pdata$`diagnosis:ch1`)] <- "Healthy"


#Expression data as a data frame
Exdata <- exprs(findata) %>% as.data.frame()


#Feature data
Fdata <- fData(findata)
Fdataother <- fData(otherplat)


#Ensure all gene IDs in fdata are identical
stopifnot(all(Fdata$ID %in% Fdataother$ID))

```


## Create a dataframe with desired annotation

A dataframe is created for the data set with Gene Symbols used to annotate the 
row names. A matching table is created so that the Gene Symbols (obtained from 
the feature data) replacing the Gene IDs is correctly matched. The sample names 
are also added as column names and the probe values for each gene are mean 
aggregated so that each gene contains one value.
The value of the first row is not mapped to a specific gene and is thus removed.

```{r}

#Match Gene symbols to ID in the correct order
matching_table <- Fdataother %>% dplyr::select(ID, Symbol)

df <- Exdata %>%
  rownames_to_column("ID") %>%
  gather(sample, expression,-ID)


df <- inner_join(matching_table, df, by="ID") %>%
  dplyr::select(-ID) %>% 
  spread(sample, expression) %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F)


#Add Gene symbols to df
df$gene <- matching_table$Symbol


#Mean aggregate the multiple probe values for same gene
df1 <- aggregate(. ~gene, data = df, mean)


#Annotate df rows (genes) and columns (samples)
row.names(df1) <- df1$gene
df1 <- df1[,-1]
colnames(df1) <- Pdata$geo_accession
df1 <- na.omit(df1)
```


## Data Visualization

The data is visualized via a pca plot to identify any underlying associations in 
the data. To change the variable that the plot is colored by, change the 
"groups" variable in the ggbiplot function. A tsne plot is also produced for
comparison to the pca results.
Since the results were seemingly well explained with a pca, the non-linear
dimensional reduction provided by a tsne could potentially be deemed 
superfluous. However, the consistent data behavior is important to note.

```{r, echo = FALSE}

#Dimension Reduction
#PCA
matdata1 <- as.matrix(df1)
pca1<- prcomp(t(matdata1), center=TRUE, scale. = TRUE)
ggbiplot(pca1,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104948")

#tSNE
gsne <- Rtsne(t(matdata1), perplexity = 10, theta = 0.01)

plot(gsne$Y, col = as.factor(Pdata$`diagnosis:ch1`), pch = 19,
     main = "tSNE plot for GSE104948")
```


The final obtained via this pipeline (variables df1, Pdata, and Fdata) are used 
in other scripts contained in this repository. 


This data is further analyzed via clustering here:
https://github.com/saezlab/VM_CKD_saezlab/blob/master/Clustering.Rmd

via DEA here:
https://github.com/saezlab/VM_CKD_saezlab/blob/master/DEA.Rmd

and via various Functional Analyses here:
https://github.com/saezlab/VM_CKD_saezlab/blob/master/Functional_Analysis.Rmd

