---
title: "Data upload"
author: Victoria Muckerson
date: 8/8/2019
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = FALSE)
```

## Summary

This document outlines the uploading of a chronic kidney disease data set GSE104948 (glomerular tissue) from GEO. The data set contains data obtained via two different platforms (GPL22945 and GPL24120); for batch effect minimization when comparing to other datasets, only the platform GPL24120 is used for analysis. This document concludes with the production of pca and tsne plots colored by a phenotypic variable.


## Download necessary libraries

```{r}

#Download libraries

library(GEOquery)
library(qusage)
library(Biobase)
library(annotate)
library(hgu133a.db)
library(ggbiplot)
library(tidyverse)
library(Rtsne)

```

## Downloading the data
The data is downloaded from GEO and the phenotypic, expression, and feature data is collected. Though only one platform is used for analysis, the feature data annotation for the data from the other platform contains necessary information that the data of interest does not. Therefore, here, the feature data for both platforms is downloaded.

```{r}

###Download the preprocessed data from GEO#######

mydata <- getGEO(GEO = 'GSE104948')
findata <- mydata[[2]]
otherplat <- mydata[[1]]


#####Separate the phenotypic, expression, and feature data from the large files#####

#Phenotypic

Pdata <- pData(findata)
Pdata$`diagnosis:ch1`[is.na(Pdata$`diagnosis:ch1`)] <- "Healthy"


#Expression data as a data frame
Exdata <- exprs(findata) %>% as.data.frame()


#Feature data
Fdata <- fData(findata)
Fdataother <- fData(otherplat)

!all(Fdata$ID %in% Fdataother$ID)

```

## Create a dataframe with desired annotation

A dataframe is created for the data set with Gene Symbols used to annotate the row names. A matching table is created so that the Gene Symbols (obtained from the feature data) replacing the Gene IDs is correctly matched. The sample names are also added as column names and the probe values for each gene are mean aggregated so that each gene contains one value.

```{r}

#Annotate data frames with Gene Symbols instead of IDs
matching_table <- Fdataother %>% dplyr::select(ID, Symbol)

df <- Exdata %>%
  rownames_to_column("ID") %>%
  gather(sample, expression,-ID)


df <- inner_join(matching_table, df, by="ID") %>%
  dplyr::select(-ID) %>% 
  spread(sample, expression) %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F)


#Adding Gene symbols to df
df$gene <- matching_table$Symbol


#Aggregate the probe values
df1 <- aggregate(. ~gene, data = df, mean)


#annotate df
row.names(df1) <- df1$gene
df1 <- df1[,-1]
colnames(df1) <- Pdata$geo_accession
df1 <- na.omit(df1)
```

## Data Visualization

The data is visualized via a pca plot to identify any underlying associations in the data. To change the variable that the plot is colored by, change the "groups" variable in the ggbiplot function. A tsne plot is also produced for comparison to the pca results.
Since the results were seemingly well explained with a pca, the non-linear dimensional reduction provided by a tsne could potentially be deemed superfluous. However, the consistent data behavior is important to note.

```{r, echo = FALSE}

######Dimension Reduction######
##PCAs

#Glomerular data set
matdata1 <- as.matrix(df1)
pca1<- prcomp(t(matdata1), center=TRUE, scale. = TRUE)
ggbiplot(pca1,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104948")

#Multidimensional scaling (mds)
gsne <- Rtsne(t(matdata1), perplexity = 10, theta = 0.01)

plot(gsne$Y, col = as.factor(Pdata$`diagnosis:ch1`), pch = 19,
     main = "tSNE plot for GSE104948")
```

The final obtained via this pipeline (variables df1, Pdata, and Fdata) are used in other scripts contained in this repository. 


This data is further analyzed via clustering here:
https://github.com/saezlab/VM_CKD_saezlab/blob/master/Clustering.Rmd

via DEA here:
https://github.com/saezlab/VM_CKD_saezlab/blob/master/DEA.Rmd

and via various Functional Analyses here:
https://github.com/saezlab/VM_CKD_saezlab/blob/master/Functional_Analysis.Rmd

