---
title: "Data upload"
author: Victoria Muckerson
date: 8/8/2019
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = FALSE)
```

## This document outlines the uploading of a chronic kidney disease data set GSE104948 (glomerular tissue) from GEO. The data set uses two platforms (GPL22945 and GPL24120) and is divided accordingly. This document concludes with the production of pca and tsne plots colored by a phenotypic variable.


## Download necessary libraries

```{r}

#Download libraries

library(GEOquery)
library(qusage)
library(Biobase)
library(annotate)
library(hgu133a.db)
library(ggbiplot)
library(tidyverse)
library(Rtsne)

```

## Downloading the data and separating the phenotypic data from the expression data. Variables named based on tissue type and platform.

```{r}

###Download the preprocessed data from GEO#######

mydata <- getGEO(GEO = 'GSE104948')
data1 <- mydata[[1]]
data2 <- mydata[[2]]


#####Separate the phenotypic, expression, and feature data from the large files#####

#Phenotypic
Pdata1 <- pData(data1)
Pdata2 <- pData(data2)


#Expression data as a data frame
Exdata1 <- as.data.frame(exprs(data1))
Exdata2 <- as.data.frame(exprs(data2))


#Feature data
fdata1 <- fData(data1)
fdata2 <- fData(data2)

```

## A dataframe is created for each data set (4) and Gene Symbols are used to annotate the row names. A matching table is created so the Gene Symbols replacing the Gene IDs are matched correctly.

##The feature data contains the symbol information and can be used for this annotation. If the feature data does not contain this information, one could use biomaRt to obtain the necessary gene annotation.

```{r}

#####Annotate data frames with Gene Symbols instead of IDs#######

##Data with Gene Symbol data
matching_table1 <- fdata1[,c("ID", "Symbol")] %>%
  as_tibble()

df1 <- Exdata1 %>%
  rownames_to_column("ID") %>%
  gather(sample, expression,-ID) %>%
  as_tibble()

df1 <- inner_join(matching_table1, df1, by="ID") %>%
  select(-ID) %>% 
  spread(sample, expression) %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F) %>%
  as.data.frame()

matching_table2 <- fdata1[,c("ID", "Symbol")] %>%
  as_tibble()

df2 <- Exdata2 %>%
  rownames_to_column("ID") %>%
  gather(sample, expression,-ID) %>%
  as_tibble()

df2 <- inner_join(matching_table2, df2, by="ID") %>%
  select(-ID) %>% 
  spread(sample, expression) %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F) %>%
  as.data.frame()


#Adding Gene symbols to df
df1$gene <- matching_table1$Symbol
df2$gene <- matching_table2$Symbol

```


##  The data frames are annotated with the Gene Symbols and sample names. If there were multiple probe values for a gene, the values were mean aggregated. The final dataframe as well as the initial Phenotypic and Expression data are saved as an rds file.


```{r}

#Aggregate the probe values
df12 <- aggregate(. ~gene, data = df1, mean)
df22 <- aggregate(. ~gene, data = df2, mean)

#annotate df
row.names(df12) <- df12$gene
df12 <- df12[,-1]
colnames(df12) <- Pdata1$geo_accession

row.names(df22) <- df22$gene
df22 <- df22[,-1]
colnames(df22) <- Pdata2$geo_accession

df12 <- na.omit(df12)
df22 <- na.omit(df22)

#Save everything
saveRDS(df12, file="./all_104948.rds")
saveRDS(Pdata1, file="./Pdata_104948.rds")
saveRDS(Exdata1, file="./Ex_104948.rds")

saveRDS(df22, file="./all_104954.rds")
saveRDS(Pdata2, file="./Pdata_104954.rds")
saveRDS(Exdata2, file="./Ex_104954.rds")

```

## The data is visualized via a pca plot to identify any underlying associations in the data. To change the variable that the plot is colored by, change the "groups" variable in the ggbiplot function.


## Pca results are saved and then a tsne plot is produced to see if results are similar. Since the results were seemingly well explained with a pca, the non-linear dimensional reduction provided by a tsne seemed superfluous and was not used.

```{r, echo = FALSE}

######Dimension Reduction######

#PCAs
matdata1 <- as.matrix(data1)
pca1<- prcomp(t(matdata1), center=TRUE, scale. = TRUE)
ggbiplot(pca1,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata1$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104948 Platform GPL22945")


matdata2 <- as.matrix(data2)
pca2<- prcomp(t(matdata2), center=TRUE, scale. = TRUE)
ggbiplot(pca2,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata2$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104948 Platform GPL24120")

#saveRDS(pca1, file="./pca_104948.rds")
#saveRDS(pca2, file="./pca_104954.rds")


#Multidimensional scaling (mds)
gsne <- Rtsne(t(matdata1), perplexity = 10, theta = 0.01)

plot(gsne$Y, col = as.factor(Pdata1$`diagnosis:ch1`), pch = 19,
     main = "tSNE plot for GSE104948 Platform 22945")
```

#This data is further analyzed via clustering here:
https://github.com/vmuckerson/Chronic_kidney_disease/blob/master/Clustering_Glom.R


# and via DEA here:
https://github.com/vmuckerson/Chronic_kidney_disease/blob/master/Diff_expression_analysis.R


