---
title: "Functional_Analysis"
author: "Victoria Muckerson"
date: "8/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

This document outlines the functional analysis of microarray datasets GSE104948
and GSE32591 and the lupus nephropathy and control samples they contain. Signal
pathway impact analysis (SPIA) is performed using the differentially expressed
genes (DEGs) between the healthy and sick phenotypes to obtain pathway
representation information. Following, Progeny is used to predict the pathway
activity.


##Load libraries

```{r}

#Load libraries

library(msigdbr)
library(SPIA)
library(tidyverse)
library(dplyr)
library(plyr)
library(progeny)

```


## Load data from previous pipelines

[data](https://github.com/saezlab/VM_CKD_saezlab/blob/master/Data_download_and_pca.Rmd)
[DEA results](https://github.com/saezlab/VM_CKD_saezlab/blob/master/DEA.Rmd)

```{r}

## Load Data
setwd("/Users/saezlab/Documents/CKD_Data/Glom_Data/")

data1 <- readRDS(file = "./all_104948.rds")
Pdata1 <- readRDS(file = "./Pdata_104948.rds")
fdata1 <- readRDS(file = "./fdata_104948.rds")

data3 <- readRDS(file = "./all_GSE32591.rds")
Pdata3 <- readRDS(file = "./Pdata_GSE32591.rds")
fdata3 <- readRDS(file = "./fdata_GSE32591.rds")

DEA_results1 <- readRDS(file="./LN_DEA_104948.rds")
DEA_results3 <- readRDS(file="./LN_DEA_32591.rds")

```


## Select DEGs
Select for "significantly" (adjusted p-value < 0.05) differentially expressed
genes between the lupus nephropathy and healthy phenotypes.

```{r}

limma_genes <- filter(DEA_results1, adj.P.Val < 0.05)[[1]]
limma_genes3 <- filter(DEA_results3, adj.P.Val < 0.05)[[1]]

DEgenes1 <- DEA_results1$Gene == limma_genes
DEG1 <- DEA_results1[DEgenes1,]

DEgenes3 <- DEA_results3$Gene == limma_genes3
DEG3 <- DEA_results3[DEgenes3,]

```


## Annotate by Entrez ID
Use the Entrez ID to annotate a data frame as the functional tools require.

```{r}

matching_table1 <- fdata1[,c("ENTREZ_GENE_ID", "Symbol")] %>%
  as_tibble() %>%
  mutate(ENTREZ_GENE_ID = as.character(ENTREZ_GENE_ID)) %>%
  dplyr::rename(Gene = Symbol)

df1 <- DEG1 %>%
  rownames_to_column("Gene") %>%
  as_tibble()

joined_df <- inner_join(matching_table1, DEG1, by="Gene") %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F) %>%
  as.data.frame()

all1 <- inner_join(matching_table1, DEA_results1, by="Gene") %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F) %>%
  as.data.frame() %>% name_rows() %>% dplyr::select(-Gene, -logFC, -AveExpr, -t, -P.Value, -adj.P.Val, -B)


matching_table3 <- fdata3[,c("ID", "Gene Symbol")] %>%
  as_tibble() %>%
  mutate(ID = as.character(ID)) %>%
  dplyr::rename(Gene = "Gene Symbol")

df3 <- DEG3 %>%
  rownames_to_column("Gene") %>%
  as_tibble()

joined_df3 <- inner_join(matching_table3, DEG3, by="Gene") %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F) %>%
  as.data.frame()

all3 <- inner_join(matching_table3, DEA_results3, by="Gene") %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F) %>%
  as.data.frame() %>% name_rows() %>% dplyr::select(-Gene, -logFC, -AveExpr, -t, -P.Value, -adj.P.Val, -B)


de1 <- joined_df %>% dplyr::select(logFC) %>% t()
names(de1) <- colnames(de1)

de3 <- joined_df3 %>% dplyr::select(logFC) %>% t()
names(de3) <- colnames(de3)

```


## Obtain reference pathways from MSigDB
Load a reference library of pathways that SPIA will use to run an ORA with the
addition of topological information.

```{r}

msig <- msigdbr(species = "Homo sapiens", category = NULL,
        subcategory = NULL)

```


## Perform SPIA
Signal pathway impact analysis is used to determine which pathways are truly
affected by the differential expression of the significant genes.

```{r}

spia_results1 <- spia(de = de1,
                      all = all1$.rownames,
                      organism = "hsa",
                      data.dir = NULL,
                      pathids = NULL,
                      nB = 2000,
                      plots = TRUE,
                      combine = "norminv")

spia_results3 <- spia(de = de3,
                      all = all3$.rownames,
                      organism = "hsa",
                      data.dir = NULL,
                      pathids = NULL,
                      nB = 2000,
                      plots = TRUE,
                      combine = "norminv")

```


## Run Progeny and perform a PCA
Progeny predicts pathway activities and returns a matrix of pathway scores that
are then used to try and separate lupus and healthy phenotypes. The result is
visualized in a pca plot.

```{r}
## Progeny


onlyLN <- Pdata1$`diagnosis:ch1` == "Healthy" | Pdata1$`diagnosis:ch1` == "Systemic Lupus Erythematosus"
data1LN <- data1[,onlyLN]
Pdata1LN <- Pdata1[onlyLN,]


pathways1 <- progeny(as.matrix(data1LN), scale = TRUE)
pathways3 <- progeny(as.matrix(data3), scale = TRUE)


pca1 <- prcomp(pathways1, center=TRUE, scale. = TRUE)
ggbiplot(pca1,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata1LN$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104948 Pathways")

pca3 <- prcomp(pathways3, center=TRUE, scale. = TRUE)
ggbiplot(pca3,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata3$disease) +
  ggtitle("PCA of GSE32591 Pathways")




```


#PCA of raw data
For visual comparison of the data, the original preprocessed data is visualized
in a pca plot.

```{r}
## PCA of raw data for comparison

pca <- prcomp(t(data1LN), center=TRUE, scale. = TRUE)
ggbiplot(pca,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata1LN$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104948")


pca <- prcomp(t(data3), center=TRUE, scale. = TRUE)
ggbiplot(pca,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata3$disease) +
  ggtitle("PCA of GSE32591")
```


#Creating a "signature"
There is an apparent improvement in the separation of healthy and lupus samples
when comparing pathway scores and the preprocessed data that can be seen in the 
increase of information explained in the first principal component. In an attempt
to improve the separation even further, a pca is used to visualize the separation
when only the consistently differentially expressed genes from the preprocessed
data is used.

```{r}
#Signature creation
overlap <- DEG1$Gene %in% DEG3$Gene
sig <- DEG1[overlap,]

overlap2 <- DEG3$Gene %in% DEG1$Gene
sig2 <- DEG3[overlap2,]

index <- rownames(data1LN) %in% sig$Gene
index3 <- rownames(data3) %in% sig2$Gene

datasig1 <- data1LN[index,]
datasig3 <- data3[index3,]


#PCA
pca <- prcomp(t(datasig1), center=TRUE, scale. = TRUE)
ggbiplot(pca,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata1LN$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104948 DEG Overlap")



pca <- prcomp(t(datasig3), center=TRUE, scale. = TRUE)
ggbiplot(pca,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata3$disease) +
  ggtitle("PCA of GSE32591 DEG Overlap")
```


#Applying signature on other datasets
In order to test robustness of the "signature", it was applied to a different
dataset with a different tissue (tubularinterstitial). Aside from these two
differences, the characteristics of the datasets (technology, disease, organ,
species, etc.) are the same. The result of the signature separation is again 
visualized in a pca with a pca of the preprocessed data following.

```{r}
alldata2 <- readRDS(file = "/Users/saezlab/Documents/CKD_Data/Tub_Data/all_104954.rds")
Pdata2 <- readRDS(file = "/Users/saezlab/Documents/CKD_Data/Tub_Data/Pdata_104954.rds")
Pdata2$`diagnosis:ch1`[is.na(Pdata2$`diagnosis:ch1`)] <- "Healthy"

onlyLN2 <- Pdata2$`diagnosis:ch1` == "Systemic lupus erythematosus" | Pdata2$`diagnosis:ch1` == "Healthy"
data2LN <- alldata2[,onlyLN2]
Pdata2LN <- Pdata2[onlyLN2,]


index2 <- rownames(data2LN) %in% sig$Gene
datasig2 <- data2LN[index2,]

pca <- prcomp(t(datasig2), center=TRUE, scale. = TRUE)
ggbiplot(pca,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata2LN$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104945 DEG Overlap")


pca <- prcomp(t(data2LN), center=TRUE, scale. = TRUE)
ggbiplot(pca,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata2LN$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104945")




try <- rownames(data3) %in% sig$Gene
leap <- data1[try,]


pca <- prcomp(t(datasig3), center=TRUE, scale. = TRUE)
ggbiplot(pca,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata3$disease) +
  ggtitle("PCA of GSE32591")

```


#DEG and Progeny
While the signature pca performed better than the preprocessed data pca with all
genes, the pathway scores outperformed both. Therefore, in order to make this
separation even more stark, the "signature" genes are used in Progeny to obtain
pathways scores. This multi-component analysis performed better than all other
methods (quantified by PC1%). This final result is seen in the following pca
plots.

```{r}
pathways1 <- progeny(as.matrix(datasig1), scale = TRUE)

pca1 <- prcomp(pathways1, center=TRUE, scale. = TRUE)
ggbiplot(pca1,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata1LN$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104948 DEG Pathways")



pathways3 <- progeny(as.matrix(datasig3), scale = TRUE)

pca3 <- prcomp(pathways3, center=TRUE, scale. = TRUE)
ggbiplot(pca1,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata3$disease) +
  ggtitle("PCA of GSE32591 DEG Pathways")


pathways2 <- progeny(as.matrix(data2LN), scale = TRUE)

pca2 <- prcomp(pathways2, center=TRUE, scale. = TRUE)
ggbiplot(pca2,
         obs.scale = 1,
         var.scale = 1,
         var.axes = FALSE,
         groups = Pdata2LN$`diagnosis:ch1`) +
  ggtitle("PCA of GSE104945 Pathways")

```


While this signature does not separate Lupus Nephropathy from other chronic
kidney diseases, it is evident that with multiscaled analysis the power of
separation and molecular disease classification increases. Therefore, it is hoped
that this study will stimulate further investigation in this field of multiscaled
analyses.

