---
title: "Functional_Analysis"
author: "Victoria Muckerson"
date: "8/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

This document outlines the functional analysis of microarray datasets GSE104948 and GSE32591 and the lupus nephropathy and control samples they contain. Signal pathway impact analysis (SPIA) is performed using the differentially expressed genes (DEGs) between the healthy and sick phenotypes to obtain pathway representation information. Following, Progeny is used to predict the pathway activity.


##Download necessary libraries

```{r}

#Download libraries

library(msigdbr)
library(SPIA)
library(tidyverse)
library(dplyr)
library(plyr)
library(progeny)

```


## Download [data] (https://github.com/saezlab/VM_CKD_saezlab/blob/master/Data_download_and_pca.Rmd) and [DEA results](https://github.com/saezlab/VM_CKD_saezlab/blob/master/DEA.Rmd)

```{r}

## Load Data
setwd("/Users/saezlab/Documents/CKD_Data/Glom_Data/")

data1 <- readRDS(file = "./all_104948.rds")
Pdata1 <- readRDS(file = "./Pdata_104948.rds")
fdata1 <- readRDS(file = "./fdata_104948.rds")

data3 <- readRDS(file = "./all_GSE32591.rds")
Pdata3 <- readRDS(file = "./Pdata_GSE32591.rds")
fdata3 <- readRDS(file = "./fdata_GSE32591.rds")

DEA_results1 <- readRDS(file="./LN_DEA_104948.rds")
DEA_results3 <- readRDS(file="./LN_DEA_32591.rds")

```


## Select DEGs

```{r}

limma_genes <- filter(DEA_results1, adj.P.Val < 0.05)[[1]]
limma_genes3 <- filter(DEA_results3, adj.P.Val < 0.05)[[1]]

DEgenes1 <- DEA_results1$Gene == limma_genes
DEG1 <- DEA_results1[DEgenes1,]

DEgenes3 <- DEA_results3$Gene == limma_genes3
DEG3 <- DEA_results3[DEgenes3,]

```


## Annotate by Entrez ID

```{r}

matching_table1 <- fdata1[,c("ENTREZ_GENE_ID", "Symbol")] %>%
  as_tibble() %>%
  mutate(ENTREZ_GENE_ID = as.character(ENTREZ_GENE_ID)) %>%
  dplyr::rename(Gene = Symbol)

df1 <- DEG1 %>%
  rownames_to_column("Gene") %>%
  as_tibble()

joined_df <- inner_join(matching_table1, DEG1, by="Gene") %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F) %>%
  as.data.frame()

all1 <- inner_join(matching_table1, DEA_results1, by="Gene") %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F) %>%
  as.data.frame() %>% name_rows() %>% dplyr::select(-Gene, -logFC, -AveExpr, -t, -P.Value, -adj.P.Val, -B)


matching_table3 <- fdata3[,c("ID", "Gene Symbol")] %>%
  as_tibble() %>%
  mutate(ID = as.character(ID)) %>%
  dplyr::rename(Gene = "Gene Symbol")

df3 <- DEG3 %>%
  rownames_to_column("Gene") %>%
  as_tibble()

joined_df3 <- inner_join(matching_table3, DEG3, by="Gene") %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F) %>%
  as.data.frame()

all3 <- inner_join(matching_table3, DEA_results3, by="Gene") %>%
  data.frame(row.names = 1, check.names = F, stringsAsFactors = F) %>%
  as.data.frame() %>% name_rows() %>% dplyr::select(-Gene, -logFC, -AveExpr, -t, -P.Value, -adj.P.Val, -B)


de1 <- joined_df %>% dplyr::select(logFC) %>% t()
names(de1) <- colnames(de1)

de3 <- joined_df3 %>% dplyr::select(logFC) %>% t()
names(de3) <- colnames(de3)

```


## Obtain reference pathways from MSigDB

```{r}

msig <- msigdbr(species = "Homo sapiens", category = NULL,
        subcategory = NULL)

```


## Perform SPIA

```{r}

spia_results1 <- spia(de = de1,
                      all = all1$.rownames,
                      organism = "hsa",
                      data.dir = NULL,
                      pathids = NULL,
                      nB = 2000,
                      plots = TRUE,
                      combine = "norminv")

spia_results3 <- spia(de = de3,
                      all = all3$.rownames,
                      organism = "hsa",
                      data.dir = NULL,
                      pathids = NULL,
                      nB = 2000,
                      plots = TRUE,
                      combine = "norminv")

```


## Run Progeny













